---
title: "analysis"
output: html_document
date: '2022-11-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = FALSE}
library(raster)
library(sf)
library(tidyverse)
library(dsm)
library(scico)
library(ggnewscale)
```

This markdown file uses corresponding RData files to run the analysis presented in "Bird, J.P., Fuller, R.A. and Shaw, J.D. 2022. Comparing the responses of extant and extirpated seabirds to the world’s largest multi-predator eradication."

It includes production of the figure components presented.

# Invasive species legacy
The current distribution and abundance of our four study species was analysed using the outputs of two supporting methods papers:

Bird, J.P., Fuller, R.A., Pascoe, P.P. and Shaw, J.D., 2022. Trialling camera traps to determine occupancy and breeding in burrowing seabirds. Remote Sensing in Ecology and Conservation, 8(2), pp.180-190.

Bird, J.P., Terauds, A., Fuller, R.A., Pascoe, P.P., Travers, T.D., McInnes, J.C., Alderman, R. and Shaw, J.D., 2022. Generating unbiased estimates of burrowing seabird populations. Ecography, p.e06204.

First, load data files.

```{r}
load("analysis_data1.RData")
load("analysis_data2.RData")
load("analysis_data3.RData")
```

## Calculating occupied area for four study species

Model-predictions from density surface models of burrow density for Antarctic Prions and White-headed Petrels were adjusted by occupacy to calculate population density. The number of pixels predicted to support >1 breeding pair was calculated.

```{r}
# ap prediction
pr_ap <- as.data.frame(ap_raster, xy = T) %>% 
         drop_na() %>% 
         mutate(pd = dsm_ap_nb3 * ap_occ, # multiply by occupancy to model pop rather than burrow density
                pc = (dsm_ap_nb3 * ap_occ)*400) 
# 400 is the area of each 20 m x 20 m pixel

# filter only pixels with >1 breeding pair predicted
pr_apf <- pr_ap %>% filter(pc > 1)

# % occupied area
tmp <- as.data.frame(ap_raster, xy = T) %>% drop_na()
(length(pr_apf$dsm_ap_nb3)/length(tmp$dsm_ap_nb3))*100
```

```{r}
# whp prediction
pr_whp <- as.data.frame(whp_raster, xy = T) %>% 
          drop_na() %>% 
          mutate(pd = dsm_whp_tw2 * whp_cocc, # multiply by occupancy to map pop rather than burrow density
                 pc = (dsm_whp_tw2 * whp_cocc)* 400)
# 400 is the area of each 20 m x 20 m pixel

# filter only pixels with >1 breeding pair predicted
pr_whpf <- pr_whp %>% filter(pc > 1) 

# % occupied area
tmp <- as.data.frame(whp_raster, xy = T) %>% drop_na()
(length(pr_whpf$dsm_whp_tw2)/length(tmp$dsm_whp_tw2))*100
```

For Blue and Grey Petrels the proportion of pixels occupied in islamd-wide population density layers from targeted search surveys was calculated.

```{r bp stats}
# of pixels in island-wide grid
gd <- 3699840
# % pixels occupied
(length(gdat_bp$dens)/gd)*100
```

```{r gp stats}
# % pixels occupied
(length(gdat_gp$dens)/gd)*100
```

## Figure 1
Figure 1 has a number of components. 

### Plot area
First, adjusting scales for plotting.

```{r adjusting lat/long scales}

scale_x_longitude <- function(xmin=-180, xmax=180, step=1, ...) {
    ewbrks <- seq(xmin,xmax,step)
    ewlbls <- unlist(lapply(ewbrks, function(x) ifelse(x < 0, paste(x, "W"), ifelse(x > 0, paste(x, "E"),x))))
    return(scale_x_continuous("Longitude", breaks = ewbrks, labels = ewlbls, expand = c(0, 0), ...))
}
scale_y_latitude <- function(ymin=-90, ymax=90, step=0.5, ...) {
    nsbrks <- seq(ymin,ymax,step)
    nslbls <- unlist(lapply(nsbrks, function(x) ifelse(x < 0, paste(x, "S"), ifelse(x > 0, paste(x, "N"),x))))
    return(scale_y_continuous("Latitude", breaks = nsbrks, labels = nslbls, expand = c(0, 0), ...))
}
```

Reading a spatial layer of the coastline and setting the coordinate reference system.

```{r}
# set CRS from coast layer
crs <- st_crs(ccoast)
# buffer coastline for neater plotting
ccoast_buff <- st_buffer(ccoast, dist = 800)
```

### Panel A - relief map
Create hillshade for the relief map.

```{r}
# create slope and hillshade
slope = terrain(dem, opt='slope')
aspect = terrain(dem, opt='aspect')
hill = hillShade(slope, aspect, 40, 270)
```

Generate the first panel - a relief map of Macquarie Island. 

```{r}
dem_df <- as.data.frame(dem, xy = T) %>% 
         drop_na()
hill_df <- as.data.frame(hill, xy = T) %>% 
         drop_na()

ggplot() +
               geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
               geom_raster(aes(x=x, y=y, fill=layer), data = hill_df) + 
               scale_fill_gradient(low = "black", high = "white") +
               new_scale_fill() + 
               geom_raster(aes(x=x, y=y, fill=Macca_dem), data = dem_df, alpha = 0.2) + 
               scale_fill_gradientn(colours = terrain.colors(10)) + 
               scale_x_longitude(158.8, 158.9, step = 0.1) +
               theme_bw()
```

### Panel B - combined species distribution map
Species densities were normalised to present all species together

```{r}
# normalise densities for all spp.
pr_apr <- pr_apf %>% filter(pd > quantile(pd, 0.025) & pd < quantile(pd, 0.975)) %>% 
         mutate(pd_norm = (pd-min(pd))/diff(range(pd)))
pr_whpr <- pr_whpf %>% filter(pd > quantile(pd, 0.025) & pd < quantile(pd, 0.975)) %>% 
           mutate(pd_norm = (pd-min(pd))/diff(range(pd)))
gdat_gpr <- gdat_gp %>% filter(dens > quantile(dens, 0.025) & dens < quantile(dens, 0.975)) %>% 
         mutate(pd_norm = (dens-min(dens))/diff(range(dens)))
gdat_bpr <- gdat_bp %>% filter(dens > quantile(dens, 0.025) & dens < quantile(dens, 0.975)) %>% 
         mutate(pd_norm = (dens-min(dens))/diff(range(dens)))

ggplot() +
    # set-up plot background etc.
    geom_sf(fill = 'grey25', data = ccoast) +
    geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
    scale_x_longitude(158.8, 158.9, step = 0.1) +
    # add GP layer #440154FF #E0754FFF
    geom_sf(data = gdat_gpr, aes(col = pd_norm), show.legend = T) +
    scale_color_gradient(low = "transparent", high = "#440154FF", aesthetics = 'colour') +
    # add BP layer #33638DFF #2A5934FF
    new_scale_colour() +
    geom_sf(data = gdat_bpr, aes(col = pd_norm), show.legend = T) +
    scale_color_gradient(low = "transparent", high = "#33638DFF", aesthetics = 'colour') +
    # add WHP layer #3CBB75FF #2B5B99FF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_whpr) +
    scale_fill_gradient(low = "transparent", high = "#3CBB75FF", aesthetics = 'fill') +
    # add AP layer #FDE725FF #C36D9AFF
    new_scale_fill() +
    geom_raster(aes(x=x, y=y, fill=pd_norm), data = pr_apr) +
    scale_fill_gradient(low = "transparent", high = "#FDE725FF", aesthetics = 'fill') +
    theme_bw()
```

### Panel C - F - comparison with 1970s distribution

Brothers (1984) surveyed the whole of Macquarie Island, determining occupancy at a 1km2 grid level and estimating abundance on a five point  scale: abundant, common, sparse, rare, absent, giving estimated quantitative thresholds for each category. 

Re-scaling the density data - Brothers reported the largest colony of White-headed Petrels was 800 pairs. His top bin for WHP is 500 - 'abundant'. It's not clear whether this is the middle of the bin, or a lower bound. Assuming it is a lower bound, I've taken 500/max(800) = ~0.6 as the density cut-off and apply that to our DSM. So, grid cells with >60% of the max cell in the DSM all get a category 4 score 'abundant' - and so on down. 

This approach tells us whether the distribution of the population has expanded or contracted - i.e. does the current population meet equivalent thresholds for 'abundant':'rare' in as many cells as it did in the 1970s. 

#### Panel C

```{r a}
#Select AP grid squares in 1970s Brothers dataset
bros_ap <- bros %>% st_centroid() %>% filter(A_Prion_Br != 0)

# Plot map
ggplot() +
  geom_sf(fill = 'grey', data = ccoast) +
  geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
  geom_sf(data = bros_ap, aes(col = A_Prion_Br)) +
  scale_color_gradient(low = "#4401541A", high = "#440154FF", aesthetics = 'colour') +
  scale_x_longitude(158.8, 158.9, step = 0.1) +
  theme_bw()
```

#### Panel D

```{r b}
# Re-sample raster layer of predicted AP density using same grid used by Brothers 
bros_comp_ap <- bros %>% mutate(ap_pred = raster::extract(ap_raster, ., fun = sum, na.rm = T)) %>% 
                st_centroid() 
bros_comp_ap <- bros_comp_ap[order(bros_comp_ap$ap_pred),] %>% mutate(cumsum = cumsum(ap_pred))

# Define quantiles for classifying predicted density
ap_q1 <- bros_q %>% filter(species=="ap", !is.na(percentile_sq))
ap_q1 <- ap_q1[[4]]

# Classify predicted density
bros_comp_ap <- bros_comp_ap %>% mutate(q1 = case_when(ap_pred > (max(ap_pred) * ap_q1[1]) ~ 4,
                                                       ap_pred > (max(ap_pred) * ap_q1[2]) ~ 3,
                                                       ap_pred > (max(ap_pred) * ap_q1[3]) ~ 2,
                                                       ap_pred > (max(ap_pred) * ap_q1[4]) ~ 1,
                                                       TRUE ~ 0))
# Plot map
ggplot(data = filter(bros_comp_ap, q1 != 0)) +
                     geom_sf(fill = 'grey', data = ccoast) +
                     geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
                     geom_sf(aes(col = q1), show.legend = T) +
                     scale_color_gradient(low = "#4401541A", high = "#440154FF",
                                          aesthetics = 'colour') +
                     scale_x_longitude(158.8, 158.9, step = 0.1) +
                     theme_bw()
```

#### Panel E

```{r c}
#Select WHP grid squares in 1970s Brothers dataset
bros_whp <- bros %>% st_centroid() %>% filter(WH_Petrel_ != 0)

# Plot map
ggplot() +
  geom_sf(fill = 'grey', data = ccoast) +
  geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
  geom_sf(data = bros_whp, aes(col = WH_Petrel_), show.legend = T) +
  scale_color_gradient(low = "#FDE7251A", high = "#FDE725FF", aesthetics = 'colour') +
  scale_x_longitude(158.8, 158.9, step = 0.1) +
  theme_bw()
```

#### Panel F

```{r WHP vs Brothers - method 1}
# Re-sample raster layer of predicted WHP density using same grid used by Brothers
bros_comp_whp <- bros %>% mutate(whp_pred = 400*(raster::extract(whp_raster, ., fun = sum, na.rm = T))) %>% 
                st_centroid()
bros_comp_whp <- bros_comp_whp[order(bros_comp_whp$whp_pred),] %>% mutate(cumsum = cumsum(whp_pred))

# Define quantiles for classifying predicted density
whp_q1 <- bros_q %>% filter(species=="whp", !is.na(percentile_sq))
whp_q1 <- whp_q1[[4]]

# Classify predicted density
bros_comp_whp <- bros_comp_whp %>% mutate(q1 = case_when(whp_pred > (max(whp_pred) * whp_q1[1]) ~ 4,
                                                       whp_pred > (max(whp_pred) * whp_q1[2]) ~ 3,
                                                       whp_pred > (max(whp_pred) * whp_q1[3]) ~ 2,
                                                       whp_pred > (max(whp_pred) * whp_q1[4]) ~ 1,
                                                       TRUE ~ 0))
# Plot map
ggplot(data = filter(bros_comp_whp, q1 != 0)) +
  geom_sf(fill = 'grey', data = ccoast) +
  geom_sf(fill = NA, colour = NA, data = ccoast_buff) +
  geom_sf(aes(col = q1), show.legend = T) +
  scale_color_gradient(low = "#FDE7251A", high = "#FDE725FF", aesthetics = 'colour') +
  scale_x_longitude(158.8, 158.9, step = 0.1) +
  theme_bw()
```


# Why do recovery trajectories differ?

## Trends from long-term monitoring data

### Figure 2

#### Panel A
Generalized Additive Model smooths of burrow numbers in monitoring plots

```{r brw ct}
# include species as a factor, individual species-level smooths of year, and site as a random effect
gam_1 <- dat %>% filter(!is.na(brw_ct)) %>% 
         gam(brw_ct ~ species + 
                      s(year, by = species, bs = "tp", m=2) + 
                      s(site, bs="re"), data = ., family = "nb")

# check gam performance
summary(gam_1)
par(mfrow = c(2,2))
gam.check(gam_1)

# predict terms for each species
pred <- data.frame(year = seq(min(1988), max(2018), length.out = 1001)) %>% 
            mutate(species = gl(4, 1, length = 1001, 
                             labels = c("AP","BP","GP","WHP")),
                   site = gl(4, 1, length = 1001, 
                             labels = c("Hill 291","Sodomy West","Douglas Pt North","Douglas Pt South"))) %>% 
            cbind(predict(gam_1, se.fit = TRUE, ., type = "terms", exclude = "s(site)"))

predfit <- pred %>% pivot_longer(cols = c(5:8), names_to = "sp", values_to = "fit") %>% 
           left_join(pred %>% pivot_longer(cols = c(10:13), names_to = "sp1", values_to = "se")) %>% 
           filter(fit != 0 & se != 0, species != "GP" | year > 2001, species != "BP" | year > 2000) %>% 
           dplyr::select(year, species, fit, se)

# plot combined marginal effects plot of trends for all species
ggplot(data = predfit, aes(year, fit)) +
  geom_line(aes(colour = species)) +
  geom_ribbon(aes(x = year, ymin = fit - se, ymax = fit + se, fill = species), alpha = 0.1) +
  scale_colour_viridis_d(aesthetics = c("colour", "fill")) +
  theme_bw()
```

#### Panel B
Generalized Additive Model smooths of breeding success in monitoring plots

```{r br succ}
# set up breeding data
brd <- dat %>% filter(!is.na(fledge)) %>% 
               group_by(species, year) %>% 
               summarise(br_attempt = sum(br_attempt),
                         fledge = sum(fledge),
                         fail = sum(fail),
                         fl_fa = sum(fl_fa)) %>% 
               mutate(br_succ = fledge/(fledge + fail),
                      br_succ_min = (fledge/br_attempt),
                      br_succ_max = ((fledge + !is.na(fl_fa))/br_attempt),
                      era = case_when(year < 1995 ~ "cats",
                                      TRUE ~ "nocats"))

# gam of breeding success by species by year. Era is pre and post cat eradication
gam_2 <- gam(cbind(fledge, fail) ~ species + era + s(year, by = species, bs = "tp", m=2), data = brd, family = "binomial", method = "REML")

# check gam performance
summary(gam_2)
gam.check(gam_2)

# use the model to predict breeding success across the whole time series
suc_pred <- data.frame(year = seq(min(1960), max(2018), length.out = 1001)) %>%
            mutate(species = gl(2, 1, length = 1001, 
                             labels = c("GP","WHP")),
                   era = case_when(year < 1996 ~ "cats",
                                   TRUE ~ "nocats")) %>% 
            cbind(predict(gam_2, se.fit = TRUE, ., type = "link"))
suc_pred$fitn <- gam_2$family$linkinv(suc_pred$fit)
suc_pred$upr <- gam_2$family$linkinv(suc_pred$fit + 2*suc_pred$se.fit)
suc_pred$lwr <- gam_2$family$linkinv(suc_pred$fit - 2*suc_pred$se.fit)
suc_pred <- suc_pred %>% filter(species != "GP" | year > 2001)

# points to include on the plot
pts <- brd %>% dplyr::select(year, fitn=br_succ) #%>% filter(year > 1975)

# plot breeding success
ggplot(suc_pred, aes(year, fitn)) +
        annotate(geom = "rect", xmin = 1996, xmax = 2000.45, ymin = 0, ymax = 1, alpha = .2) +
        annotate(geom = "rect", xmin = 2000.4, xmax = 2000.5, ymin = 0, ymax = 1) +
        annotate(geom = "rect", xmin = 2004, xmax = 2010.45, ymin = 0, ymax = 1, alpha = .2) +
        annotate(geom = "rect", xmin = 2010.45, xmax = 2011.45, ymin = 0, ymax = 1, alpha = .8) +
        geom_ribbon(aes(ymin=lwr,ymax=upr, fill = species), alpha=.1) +
        #geom_line(aes(year, lwr), linetype = "dotted") +
        #geom_line(aes(year, upr), linetype = "dotted") +
        geom_point(data=pts, aes(x=year, y=fitn, colour = species)) +
        geom_line(aes(colour = species)) +
        scale_colour_discrete(type = c("#35B779FF", "#FDE725FF"), aesthetics = c("colour", "fill")) +
        #geom_line(data=pts, aes(x=year, y=fitn), linetype = "dotted", alpha=.5) +
        xlab("Year") +
        ylab("Breeding success") +
        theme_bw()
```

### Lambda
Population growth rates from the model were calculated in excel. See lambda.xlsx

### Figure 3
Density plots of environmental features used by each species. Environmental data are extracted from a raster stack containing layers for elevation, ndvi, aspect, slope, ridge, topographic wetness and windiness on Macquarie Island.

```{r}
# Antarctic Prion density and environmental values
hab_ap <- seg_ap %>% mutate(dens = predict(dsm_ap_nb3, type = "response"),
                            aob_d = dens * ap_occ, 
                            sp = "Antarctic Prion") %>% 
          pivot_longer(cols = c(dem, ndvi, aspect, slope, ridge, wetness, wind), names_to = "var",
                       values_to = "val") %>% 
          dplyr::select(sp, aob_d, var, val)
```

```{r}
# White-headed Petrel density and environmental values
hab_whp <- seg_whp %>% mutate(dens = predict(dsm_whp_tw2, type = "response"),
                            aob_d = dens * whp_cocc,
                            sp = "White-headed Petrel") %>% 
           pivot_longer(cols = c(dem, ndvi, aspect, slope, ridge, wetness, wind), names_to = "var",
                       values_to = "val") %>% 
           dplyr::select(sp, aob_d, var, val)
```

```{r}
bp_gdot <- st_centroid(gdat_bp)

# Blue Petrel density and environmental values
hab_bp <- as.data.frame(raster::extract(sta, bp_gdot)) %>% bind_cols(bp_gdot) %>% 
          dplyr::select(dens, dem, ndvi, slope, aspect, ridge, wetness, wind) %>% 
          pivot_longer(cols = -dens, names_to = "var",
                       values_to = "val") %>% 
          mutate(sp = "Blue Petrel") %>% 
          dplyr::select(sp, aob_d = dens, var, val)
```

```{r}
gp_gdot <- st_centroid(gdat_gp)

# Grey Petrel density and environmental values
hab_gp <- as.data.frame(raster::extract(sta, gp_gdot)) %>% bind_cols(gp_gdot) %>% 
          dplyr::select(dens, dem, ndvi, slope, aspect, ridge, wetness, wind) %>% 
          pivot_longer(cols = -dens, names_to = "var",
                       values_to = "val") %>% 
          mutate(sp = "Grey Petrel") %>% 
          dplyr::select(sp, aob_d = dens, var, val)
```

```{r}
# Combined environmental values for all species
hab <- bind_rows(hab_ap, hab_whp, hab_bp, hab_gp)

ggplot() +
  geom_density(data = hab, aes(val, group = sp, weight=aob_d, fill = sp), alpha = 0.3) +
  facet_wrap(~var, scale = "free") +
  scale_fill_viridis_d() +
  theme_bw()
```

